---
title: "Predict your age from you name, and other such questions"
author: "Jeffrey B. Arnold"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Statement of the problem

Suppose I know your sex and name, can I guess you age?

The probability of an age given a name and sex,
$$
P(\text{age} | \text{name}, \text{sex}) \propto P(\text{name} | \text{age}, \text{sex}) P(\text{age} | \text{sex})
$$


## Data Wrangling

The source of our data is the **babynames** package in R.

```{r}
YEAR <- 2015
```
We'll consider a single year in our analysis `r YEAR`, whcih is the last year in the `babynames` package.

The lifetables are only provided for decades 1900, ..., 2010. We need to fill in the full lifetable for all birthyears.
For years between 1900 and 2010, we will linearly impute the probability of being alive at each age for non-decadal birth-years.
We'll use 2010 for all birth-years after 2010.
```{r life_table}
life_table <- babynames::lifetables %>%
  group_by(year) %>%
  mutate(px = lx / 1e+05) %>%
  rename(age = x) %>%
  select(year, age, sex, p_alive = px) %>%
  ungroup() %>%
  complete(sex, year = seq(min(year), YEAR, by = 1), age) %>%
  rename(birthyear = year) %>%
  group_by(sex, age) %>%
  arrange(sex, age, birthyear) %>%
  mutate(p_alive = zoo::na.approx(p_alive, na.rm = FALSE)) %>%
  fill(p_alive, .direction = "down") %>%
  ungroup() %>%
  arrange(sex, age, birthyear)
```

For this analysis, we only need the age distribution in `r YEAR`.
```{r age_distr}
age_distr <- life_table %>%
  mutate(year = birthyear + age, 
         # convert sex to character to avoid join warning
         sex = as.character(sex)) %>%
  filter(year == YEAR) %>%
  select(-year)
```
```{r}
glimpse(lifetable_full)
```
However, the `p_alive` column of `age_distr` only provides the probability of being alive in `r YEAR` conditional on having been born in a given year (and sex),
```{r fig.cap="Probability of "}
ggplot(age_distr,
       aes(x = birthyear, y = p_alive, color = sex)) +
  geom_point() +
  geom_line() +
  labs(y = expression(paste("P", group("(", paste(alive, "|", year), ")"))),
       x = "Year of Birth")
```
We need the number born each year to be able to calculate the number alive in each year, and the age distribution. 
Suppose that the number born in each year was equal, the age distribution would be:
```{r}
age_distr %>%
  group_by(sex) %>%
  mutate(p = p_alive / sum(p_alive)) %>%
ggplot(aes(x = age, y = p, color = sex)) +
  geom_point() +
  geom_line() +
  labs(x = "Age", y = "P(age)") +
  theme(legend.pos = "bottom")
```

As a proxy for the number born each year we'll use the proportion of Social Security applicants each year, provided by the `babynames::applicants`.
Since the baby-name information will also come from the Social Security data this is no less restrictive.
```{r fig.caption="Number of Social Security Applicants per year."}
ggplot(babynames::applicants, aes(x = year, y = n_all / 1e6, color = sex)) +
  geom_point() +
  geom_line() +
  labs(x = "", y = "SSA Applicants (mn)") +
  theme(legend.pos = "bottom")  
```
Clearly, the number of births is not constant per year.

Join the SSA applicant numbers and calculate the probability of each age by sex in `r YEAR`.
```{r age_distr_2}
age_distr <- left_join(age_distr,
                       rename(babynames::applicants, n_apps = n_all),
                       by = c("sex", "birthyear" = "year")) %>%
  mutate(n_alive = p_alive * n_apps) %>%
  group_by(sex) %>%
  mutate(p_age = n_alive / sum(n_alive)) %>%
  ungroup() %>%
  arrange(sex, age)
                 
```

After accounting for different numbers of births in each year, the
age distribution is different.
```{r}
ggplot(age_distr, aes(x = age, y = p_age, color = sex)) +
  geom_point() +
  geom_line() +
  labs(x = "Age", y = "P(age)") +
  theme(legend.pos = "bottom")
```

The `babynames` dataset has the number in each sex born each year with a given name (and registered by the SSA).
```{r}
glimpse(babynames)
```
The column `prop` is the proportion of people of that gender with that name born in each year, $P(\text{name} | \text{age}, \text{sex})$.

```{r babynames}
baby_names <- babynames::babynames
```
Also, since the SSA only releases names with > 5 individuals in a year, add
an additional entry for each year for babies born and given rare names.
```{r babynames_other}
babynames_other <- baby_names %>%
  group_by(sex, year) %>%
  summarise(p_named = sum(prop), n_named = sum(n)) %>%
  mutate(prop = 1 - p_named, n = n_named / p_named * prop,
         name = "OTHER")
```
```{r babynames2}
baby_names <- bind_rows(baby_names, babynames_other)
```


## Proabability of age given name and sex

Consider someone with the name of "Emma" and sex is "female".
What is the posterior distribution of their age,
$$
p(\text{age} | \text{name} = \text{"Emma"}, \text{sex} = \text{"F"}) = 
p(\text{name} = \text{"Emma"} | \text{age}, \text{sex} = \text{"F"}) p(\text{age} | \text{sex} = \text{"F"}) .
$$


```{r}
name <- "Emma"
sex <- "F"
```
Filter `babynames` to only include observations for the name "`r name`" and sex "`r name`":
```{r p_name_age}
p_name_age <- baby_names %>%
  filter(name == !!name, sex == !!sex) %>%
  select(-sex, -name) %>%
  mutate(age = YEAR - year)
```

```{r plot_name_age,fig.caption=glue("Proportion of babies in year with name of {name} and sex {sex}."}
ggplot(p_name_age, aes(x = year, y = prop)) +
  geom_line() +
  labs(x = "", y = "Proportion births")
```
The popularity of the name Emma first declined, then increased.
However, very few of those born when Emma was first popular are likely to still be alive.

```{r posterior}
posterior <-
  left_join(p_name_age, 
            select(filter(age_distr, sex == !!sex), birthyear, prior = p_age),
            by = c(year = "birthyear")) %>%
  rename(likelihood = prop) %>%
  # fill in missing values with 0
  mutate(prior = if_else(is.na(prior), 0, prior),
         # postrior
         post = likelihood * prior,
         # normalize posterior to sum to 1
         post = post / sum(post))
```

Let's plot the prior ($P(age)$), likelihood ($P(name | age)$),  and posterior ($P(name | age)$).
```{r plot_posterior}
posterior %>%
  select(age, post, likelihood, prior) %>%
  gather(variable, value, -age) %>%
  mutate(variable = recode(variable, post = "P(age | name)",
                           likelihood = "P(name | age)", prior = "P(age)"),
         variable = factor(variable, 
                           levels = c("P(age)", "P(name | age)", 
                                      "P(age | name)"))) %>%
  ggplot(aes(y = value, x = age)) +
  geom_line() +
  facet_wrap(~ variable, ncol = 1, scales = "free_y")

```

Alternatively, instead of calculating $p(age | name, sex)$ from Bayes' Theorem,
we can calculate it directly from the joint distribution of name, age, and sex.
$$
p(age | name, sex) = p(age, name, sex) / p(name, sex)
$$


Now calculate the probability of 
```{r}
baby_names_joint <- baby_names %>%
  # add probability alive
  left_join(select(age_distr, age, sex, birthyear, p_alive), 
            by = c("sex", year = "birthyear")) %>%
  # update probabilities, 
  mutate(prop = p_alive * prop,
         prop = if_else(is.na(prop), 0, prop),
         n = n * p_alive) %>%
  select(-n_alive, -p_alive) %>%
```

```{r}
baby_names_joint %>% 
  group_by()
```



### Questions

1. Consider a rare name - what is the probability on days in years in which there is no sample. How would you alter it?
2. Consider what would happen if sex was unknown? 
3. Which name provides the most information about a person's age?
4. Consider a name like "Jordan", "Jessie", "Dakota", "Marion"


### References

The [How to Tell Someone’s Age When All You Know Is Her Name](https://fivethirtyeight.com/features/how-to-tell-someones-age-when-all-you-know-is-her-name/) is 


## Information Theory

The *information* or *surprisal* of an outcome of random variable is the log inverse of its probability of occuring:
$$
I(X = x) = -\log(\Pr(X = x)) = \log \left( \frac{1}{\Pr(X = x)} \right)
$$
The more surprising an event is, the more information it contains.

The *entropy* of a discrete random variable ($X$) is the expected information of the random variable.
Suppose $X$ takes values $x_1, ..., x_n$, then its entropy is,
$$
H(X) = E_X(I(X)) = E_X(-\log(\Pr(X))) = - \sum_{i = 1}^n \Pr(X = x_i) \log \Pr(X = X_i)
$$

Examples. The entropy of a Bernoulli distribution with $p = 0.5$ is
```{r}
x <- c(0.5, 0.5)
-sum(x * log(x))
```
*Q:* Will the entropy of Bernoulli distribution with $p = 0.1$ be lower or higher than when $p = 0.5$?
```{r}
x <- c(0.9, 0.1)
-sum(x * log(x))
```
It is less, because there is less information when $p = 0.9$.
For a discrete distribution, the uniform distribution maximizes the entropy.

The *mutual information* of two discrete random variables $X$ and $Y$ is
$$
I(X; Y) = \sum_{y \in Y} \sum_{x \in X} p_{XY}(x, y)(x, y) \log \left( \frac{p_{XY}(x, y)}{p_X(x) p_Y(y)} \right)
$$
where $p_X$, $p_Y$ are the marginal distributions of $X$ and $Y$ respectively.

Suppose $X$ and $Y$ are independent? What is the mutual independence of $X$ and $Y$?
$$
\log\left( \frac{p_{XY}(x, y)}{p_X(x) p_Y(y)} \right) = \log\left( \frac{p_{X}(x) p_{Y}(y)}{p_X(x) p_Y(y)} \right) =  \log 1 = 0
$$
where the second term comes from independence.

- non-negative: $I(X; Y) \geq 0$
- symmetric: $I(X; Y) \geq I(Y; X)$
- $I(X; Y) = 0$ if and only if $X$ and $Y$ are independent random variables
- $I(X; X) = H(X)$


Kullback-Leibler divergence is used to measure the "distance" between distribution.
The divergence between probability distributions $P$ and $Q$ is
$$
D_{KL}(P(i) || Q(i)) = - \sum_i P(i) \log \left(\frac{Q(i)}{P(i)}\right) = \sum_i P(i) \log \left( \frac{P(i)}{Q(i)} \right)
$$
Note that the KL divergence is not symmetric,
$$
D_{KL}(P || Q) \neq D_{KL}(Q || P) ,
$$
and thus not a distance metric.


Some properties of KL divergence

- non-negative: $D_{KL}(P || Q) \geq 0$. Also $D_{KL}(P || Q)$ if and only if $P = Q$ almost everywhere.
- invariant to parameter transformations


