# Model Selection

We want to model $y = (y_1, \dots, y_n)$, which is a vector of $n$ non-negative integers.
For each observation $i$, we also have a $p \times 1$ vector of predictors, $x_i$.

Poisson Model
$$
\begin{aligned}[t]
y_i &\sim \mathsf{Poisson}(\lambda_i) && y_i \in \{0, 1, 2, \dots \}, i \in 1, \dots, n \\
\lambda_i &\sim \exp(\eta_i)  && \lambda_i \in \mathbb{R}^{+}, i \in 1, \dots, n \\
\eta_i &= \alpha + \beta x'_i && \eta_i \in \mathbb{R}, i \in 1, \dots, n \\
\alpha &\sim \mathsf{Normal}(0, 10) && \alpha \in \mathbb{R} \\
\beta_j &\sim \mathsf{Normal}(0, 2.5) && \beta_j \in \mathbb{R}, j \in 1, \dots, p
\end{aligned}
$$
where $x_i$ is a $k \times 1$ vector of covariates.
For the prior scales of the regression intercept $\alpha$ and coefficients, $\beta_k$, are
the weakly informative priors suggested by Stan.

Negative Binomial Model

$$
\begin{aligned}[t]
y_i &\sim \mathsf{NegBinomial2}(\mu_i, \phi) && y_i \in \{0, 1, 2, \dots \}, i \in 1, \dots, n \\
\mu_i &\sim \exp(\eta_i)  && \lambda_i \in \mathbb{R}^{+}, i \in 1, \dots, n \\
\eta_i &= \alpha + x'_i \beta && \eta_i \in \mathbb{R}, i \in 1, \dots, n \\
\alpha &\sim \mathsf{Normal}(0, 10) && \alpha \in \mathbb{R} \\
\beta_j &\sim \mathsf{Normal}(0, 2.5) && \beta_j \in \mathbb{R}, j \in 1, \dots, p \\
\phi^{-1/2} &\sim \mathsf{Exponential}(0, 1) && \phi \in \mathbb{R}^{+}
\end{aligned}
$$
where $x_i$ is a $k \times 1$ vector of covariates.
The parameter $\phi$ controls the overdispersion.
In the Poisson distribution, the variance always equals the mean, in the
negative binomial distribution, the variance is always greater than the mean.
As $\phi \to \infty$, $\mathsf{NegBinomial2(\mu, \phi)} \to \mathsf{Poisson}(\mu)$, and
as $\phi \to 0$, the overdispersion increases.
Since the less complex model is one without dispersion, we assign a prior to $\phi^{-1/2}$.
This is means that as $\phi^{-1/2} \to 0$, there is less overdispersion.
The prior is a weakly informative one.

Zero-Inflation Poisson

A common occurrence in count data is an over-abundance of 0 values relative to that predicted by a Poisson or negative binomial models.
A zero inflation model is model for this phenomena.
 See Ch. 13.7 "Zero-Inflated and Hurdle Models " in the Stan Modeling Language Reference (v 2.17).

The zero inflation Poisson model, models $y_i$ as coming from two processes.
With probability $\theta$, it comes from a process generating the excess 0s, and with 
probability $1 - \theta$, it comes from a Poisson process.
The likelihood of this model then becomes,
$$
\begin{aligned}[t]
p(y_i | \theta, \lambda_i) &= \begin{cases}
\theta + (1 - \theta) \times \mathsf{Poisson}(\lambda_i) & \text{if } y_i = 0 \\
(1 - \theta) \times \mathsf{Poisson}(\lambda_i) & \text{if } y_i > 0
\end{cases}  && y_i \in \{0, 1, 2, \dots \}, i \in 1, \dots, n 
\end{aligned}
$$
The rest of the model is the same as the Poisson model.

The zero inflated negative binomial model is similarly defined.
The likelihood of the zero-inflated negative binomial model is
$$
\begin{aligned}[t]
p(y_i | \theta, \lambda_i) &= \begin{cases}
\theta + (1 - \theta) \times \mathsf{NegBinomial2}(\mu_i, \phi) & \text{if } y_i = 0 \\
(1 - \theta) \times \mathsf{NegBinomial2}(\mu_i, \phi) & \text{if } y_i > 0
\end{cases}  && y_i \in \{0, 1, 2, \dots \}, i \in 1, \dots, n 
\end{aligned}
$$
The rest of the model is the same as the negative binomial model.

```{r}

```


1.  Some Stan questions:

    1.   What is the difference between the `neg_binomial` and `neg_binomial_2`
         disributions?
         
    1.   Why does the code use `neg_binomial_2_log` and `poisson_log`? 
         How would you write the models using the `neg_binomial_2` and `poisson`
         distributions? 

1.  The `zinfl_neg_binomial.stan` and `zinfl_poisson.stan` models define
    functions to calcualte the log probability mass of the zero inflated poisson
    and negative binomial functions.

    ```
      real zinfl_neg_binomial_2_log_lpmf(int y, real theta, real eta, real phi) {
        if (y == 0) {
          return log_sum_exp(bernoulli_lpmf(1 | theta),
                                bernoulli_lpmf(0 | theta) +
                                neg_binomial_2_log_lpmf(y | eta, phi));
        } else {
          return bernoulli_lpmf(0 | theta) +
            neg_binomial_2_log_lpmf(y | eta, phi);
        }
      }
    ```

    1.  What is the variable type that `zinfl_poisson_log_lpmf` returns?
        What is the variable type that `zinfl_poisson_rng` returns? 
        Are they the same? If they are different, why?
        
    1.  What does the function `log_sum_exp` do? How is it different than the following? 
        ```
        return bernoulli_lpmf(1 | theta) + bernoulli_lpmf(0 | theta) * poisson_log_lpmf(y | eta, phi))
        ```
        
    1.  Why does the function name end in `_lpmf`? (See Ch 24.5)? 
    
    1.  Later `zinfl_poisson_log_pmf` function is called as
        ```
        y[n] ~ zinfl_poisson_log(eta[n]);
        ```
        This is equivalent to 
        ```
        target += zinfl_poisson_log_lpmf(y[n], eta[n]);
        ```
        What is the statement `target +=` doing? 
        How does this relate to the the underlying MCMC algorithm that Stan uses?
                                
1.  For the Poisson model, use posterior predictive checks to check for

    1.  Zero-inflation
    2.  Overdispersion (variance is greater than the mean)
    
    Think of your own ways to check these quantities using the output of the 
    model and `rstan` and `bayesplot`.
    
1.  Calculate the expected log predictive density (elpd) using PSIS-LOO, using
    the **loo** package and `loo` function.

    1.  Which model has the highest elpd?
    1.  Did `loo` produce any error or warning messages? What do they mean and how would you fix them? 
    1.  The default print method for `loo` objects reports a `elpd_loo`, `p_loo`, 
        and `looic`. What is the difference between `elpd_loo` and `looic`?
        
    1.  The default print method for `loo` objects reports `p_loo`, the effective number of parameters in the model.
        It is reported with a standard error, suggesting it is an estimate.
        Why would the number of parameters in a Bayesian model need to be estimated?
        
1.  Using the **loo** package, calculate the model weights to use for Bayesian model averaging. 
    
    
    1.  Which model is given the most weight? 
    1.  Calculate the posterior predictive density from the BMA model.

1.  It is preferrable to use a continuous model expansion rather than model averaging,
    if it is possible possible. Could you incorporate these models into a single model?
    Write down and estimate that model.
    
## Model Averaging Example





    
## Shrinkage Estimators





## Mixture of Normal Distributions

A scale mixture of normal distribution is a distribution which can be written as a normal distribution conditional on the scale.
In a scale-mixture of normal distributions, a random variable $Y$ is distributed $p_Y$ is the marginal distribution of a normal distribution conditional on its scale, $W$, which distributed $p_W$,
$$
p_Y(y) = \int_{0}^\infty \mathsf{Normal}(y | 0, w) p_W(w) dw
$$
This can alo be written as,
$$
\begin{aligned}[t]
y|w &\sim \mathsf{Normal}(0, w) \\
w &\sim p_W(.) 
\end{aligned}
$$
Common examples of scale-mixtures of normal distributions include the Student-t distribution and the Laplace (Double Exponential) distribution.
The normal distribution can be thought of as a trivial scale mixture of itself, where the prior distribution of $w$ is a point mass on the scale of the normal distribution.
Also, there can be discrete scale mixtures of normal distributions,
$$
p_Y(y) = \sum_{k = 1}^K \pi_k \mathsf{Normal}(y | 0, w_k)
$$
where $pi_k \geq 0$, and $\sum_{k = 1}^K \pi_k = 1$.

A useful form of scale-mixtures of normal distributions for regularization and shrinkage is the local-global scale-mixture of normal distributions.
Each observation is independently distributed normal conditional on a global scale term ($\tau$) and a observation-specific local scale paramter ($\lambda_i$),
$$
y_i | \tau, \lambda_i \sim \mathsf{Normal}(0, \tau \lambda_i) .
$$
The prior distribution given to $\lambda_i$ determines the shrinkage properties of the marginal distribution.
The global scale parameter $\tau$ determines the overall spread of the errors.

This exercise will provide you some intuition about these distributions.

The data frame `randsamps` contains samples from the following local-global scale-mixture of normal distributions. 

| Column | Distribution of $\lambda_i$ |
|-------|--------------|
|   `x` | $\lambda_i = 1$ for all $i$                    | 
|  `y1` | $\lambda_i^{-2} \sim \mathsf{Gamma}(1/2, 1/2)$ |
|  `y2` | $\lambda_i \sim \mathsf{Cauchy}^{+}(1)$        |
|  `y3` | $\lambda_i^{-2} \sim \mathsf{Gamma}(2, 2)$     |
|  `y4` | $$\begin{aligned}\lambda_i | z_i = \begin{aligned}[t] \lambda_i | z_i = 1 \sim \mathsf{Normal}(0, 10^{-6}) \\ \lambda_i | z_i = 0 \sim \mathsf{Normal}(0, \sqrt{2}) \\ z_i &\sim \mathsf{Bernoulli}(1/2)$$ |
|  `y5`  | $\lambda_i^2 \sim \mathsf{Exponential}(1/2)$     |

```{r}
n <- 2 ^ 13
randsamps <- tibble(
  x = rnorm(n),
  y1 = x * sample(c(1e-16, sqrt(2)), size = length(x), replace = TRUE),
  y2 = x / sqrt(rgamma(length(x), 0.5, 0.5)),
  y3 = x * abs(rcauchy(length(x))),
  y4 = x / sqrt(rgamma(length(x), 2, 2)),
  y5 = x * sqrt(exp(0.5)),
)
```

The data frame `densities` contains densities for several probability density functions (with specific choices of parameters) for a range around zero.
Each probability density function matches one of the random samples 

| Column         | Description |
|----------------|------------------------------------|
| `normal`       | $\mathsf{DoubleExponential}(0, 1)$ |
| `laplace`      | $\mathsf{Laplace}(0, 1)$           |
| `student_t`    | $\mathsf{StudentT}(4, 0, 1)$       |
| `cauchy`       | $\mathsf{Cauchy}(0, 1)$            |
| `horseshoe`    | Horseshoe prior                    |
| `spike_slab`   | Spike-and-slab                     |

The Horseshoe prior does not have an analytic form, but can be approximated with,
$$
\frac{1}{\pi^2 \sqrt{2 \pi}} \log(1 + 4 / \theta^2) < p_HS (\theta) \leq  \frac{1}{\pi^2 |\theta|} .
$$

The probaibility density function of this spike-and-slab distribution is,
$$
p(y_i) = \theta \mathsf{Normal}(0, 10^-6) + (1 - \theta) \mathsf{Normal}(0, \sqrt{2}) ,
$$
where $\theta = 1/2$.

```{r}
densities <- tibble(
  x = seq(-5, 5, length.out = 2 ^ 7),
  cauchy = dcauchy(x),
  normal = dnorm(x),
  laplace = LaplacesDemon::dlaplace(x),
  student_t = dt(x, 4),
  spike_slab = 0.5 * dnorm(x, 0, 1e-6) + 0.5 * dnorm(x, 0, 100)
)
```

1.  Match the samples `x`, `y1`, ..., `y5` in `randsamples` to the density functions in  `cauchy`, `horseshoe`, `laplace`, `normal`, `spike_slab`, `spike_slab`, and `student_t`.
1.  Which of these probability density functions put the most prior probability on sparsity (values very close to zero).
        
