---
title: "Ideal Point"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Ideal Points

An ideal point (in political science) or item-response (in education) model is a latent state model.

In these models, the only observed data are votes, and the analyst wants to 
model those votes as a function of legislator- ($\theta_i$), and vote-specific ($\alpha_i$, $\lambda_i$) parameters.
The vote of legislator $i$ on roll-call $j$ ($y_{i,j}$) is a function of a
the legislator's ideal point ($\theta_i$), the vote's cutpoint ($\alpha_j$), 
and the vote's discrimination ($\lambda_j$):
$$
\begin{aligned}[t]
y_{i,j} &\sim \mathsf{Binomial}(\pi_i) \\
\pi_i &= \frac{1}{1 + \exp(-\mu_i)} \\
\mu_{i,j} &= \alpha_j + \lambda_j \theta_i
\end{aligned}
$$

**TODO** More description of these terms, and also discuss with the IRT / test example.

Ideal points (like many latent space models) are unidentified. In particular, 
there are three types of invariance:

1. Additive Aliasing
2. Multiplicative Ailising
3. Rotation (Reflection) Invariance


# Example US: Congressional Voting

```{r}
library("tidyverse")
library("pscl")
library("rstan")
```

This example models the ideology of the 109th U.S. Senate. 
This data is conveniently included in the `pscl` package:
```{r}
data("s109", package = "pscl")
```
This is not a data frame, see the documentation for its structure.

In our ideal point model, we will use binary responses, but the `s109` data
includes multiple [codes](http://voteview.com/senate109.htm) for respones to 
roll-calls.

---- ----------------------------------------------------------
0    not a member
1    Yea
2    Paired Yea
3    Announced Yea,
4    Announced Nay
5    Paired Nay
6    Nay
7    Present (some Congresses, also not used some Congresses)
8    Present (some Congresses, also not used some Congresses)
6    Nay
9    Not Voting
---- ----------------------------------------------------------

To simplify the analysis, the outcomes will be aggregated to "Yes", "No", and missing values (which will be dropped):
```{r}
votes <- s109$votes %>%
  as.data.frame() %>%
  rownames_to_column("legislator") %>%
  gather(rollcall, vote, -legislator) %>%
  mutate(y = NA,
         y = if_else(vote %in% c(1, 2, 3), TRUE, y),
         y = if_else(vote %in% c(4, 5, 6), FALSE, y),
         legislator = factor(legislator),
         rollcall = factor(rollcall, levels = colnames(s109$votes))) %>%
  filter(!is.na(y))
```

```{r}
mod_ideal_point <- stan_model("stan/ideal_point.stan")
```
```{r results='asis'}
mod_ideal_point
```

Identification of latent state models can be challenging.
Since we \textit{a priori} know, or expect, that the primary ideological dimension is Liberal-Conservative.
$$
\begin{aligned}[t]
\theta[\text{FRIST (R TN)}] & = 1 \\
\theta[\text{REID (D NV)}] & = -1
\end{aligned}
$$
Otherwise, give a weakly informative prior to the ideal points, and item difficulty and discrimination parameters,
$$
\begin{aligned}[t]
\theta_{k} &\sim \mathsf{Normal}(0, 2.5) \\
\alpha_{k} &\sim \mathsf{Normal}(0, 10.) \\
\lambda_{k} &\sim \mathsf{Normal}(0, 2.5.) 
\end{aligned}
$$


```{r}
theta <- data_frame(
  legislator = levels(votes$legislator),
  value = NA_real_
  ) %>%
  mutate(
    value = if_else(legislator == "FRIST (R TN)", -1, value),
    value = if_else(legislator == "REID (D NV)", 1, value)    
  )
```

```{r}
data_ideal_point <-
  within(list(), {
    y <- as.integer(votes$y)
    y_idx_leg <- as.integer(votes$legislator)
    y_idx_vote <- as.integer(votes$rollcall)
    Y_obs <- length(y)
    N <- nlevels(votes$legislator)
    K <- nlevels(votes$rollcall)
    # values of known ideal points
    theta_obs_idx <- which(!is.na(theta$value))
    N_obs <- length(theta_obs_idx)
    theta_param_idx <- which(is.na(theta$value))
    N_param <- length(theta_param_idx)
    theta_obs <- theta$value[theta_obs_idx]
  })
```

```{r}
sampling(mod_ideal_point, data = data_ideal_point)
```

## Identification

In ideal point model with $d$ dimensions, need $d(d + 1)$ linearly independent *a priori* restrictions:

For example, in two dimensions

- Fix ideal points of three legislators
- Fix loadings of three bills

Some references on methods to identify (Bayesian) factor, IRT, and ideal point models:

- @EroshevaCurtis2017a for a thorough discussion of identification in Bayesian factor models
- @BaiWang2014a "Identification and Bayesian Estimation of Dynamic Factor Models"
- @Quinn2004a
- Rivers, Douglas. 2003. ``Identification of Multidimensional Spatial Voting Models.'' Technical Report: Political Methodology Working Papers.

A common identification assumption is an upper triangular matrix on the loadings [@AguilarWest2000a]:
$$
\begin{bmatrix}
1             & 0             &         \dots  & 0 \\
\lambda_{2,1} & 1             &         \dots  & 0 \\
\vdots        & \vdots        &         \ddots & \vdots \\
\lambda_{k,1} & \lambda_{k,2} &         \dots  & 1 \\
\vdots        & \vdots        &         \vdots & \vdots \\
\lambda_{n,1} & \lambda_{n,2} &          \dots & \lambda_{n,k}
\end{bmatrix}
$$


