# Introduction to Stan and Linear Regression

This chapter is an introduction to writing and running a Stan model in R.
Also see the **rstan**
[vignette](https://cran.r-project.org/web/packages/rstan/vignettes/rstan.html)
for similar content.

## Prerequisites {-}

```{r setup,message=FALSE}
library("rstan")
library("tidyverse")
```

For this section we will use the `duncan` dataset included in the **carData** package.
Duncan's occupational prestige data is an example dataset used throughout the popular Fox regression text, *Applied Regression Analysis and Generalized Linear Models* [@Fox2016a].
It is originally from @Duncan1961a consists of survey data on the prestige of occupations in the US in 1950, and several predictors: type of occupation, income, and education of that
```{r}
data("Duncan", package = "carData")
```

## OLS and MLE Linear Regression

The first step in running a Stan model is defining the Bayesian statistical model that will be used for inference.

We will model `prestige` of each occupation as a function of its `education`, `occupation`, and `type`.

A standard way to do this is with the OLS estimator:
$$
\begin{multline}
y_i = \beta_0 + \beta_1 I(\mathtt{type} = \mathtt{"prof"}) + \beta_2 I(\mathtt{type} = \mathtt{"wc"}) \\
\quad + \beta_3 \mathtt{income} + \beta_4 \mathtt{education} + \epsilon_i
\end{multline}
$$

```{r}
duncan_lm <- lm(prestige ~ type + income + education, data = Duncan)
```

$$
y_i = x_i' \beta + \epsilon_i
$$
OLS finds $\hat{\beta}_{OLS}$ by minimizing the squared errors,
$$
\hat{\beta}_{\text{OLS}} = \arg \min_{b} \sum_{i = 1}^n (y_i - x_i' b)^2 .
$$
OLS is an estimator of the (linear approximation of) the conditional expectation function,
$$
\mathrm{CEF}(y_i | x_i) = E(y_i, x_i' \beta) .
$$

For valid inference we need to make assumptions about $\epsilon_i$, namely that they are uncorrelated with $X$, $\Cov(\epsilon, X) = 0$, and that they are i.i.d, $\Cov(\epsilon_i, \epsilon_j) = 0$, $\Var(\epsilon_i) = \sigma^2$ for all $i$.
However, no specific distributional form is or needs to be assumed for $\epsilon$ since CLT results show that, asymptotically the sampling distribution of $\beta$ approaches the normal.
Additionally, although $\hat\sigma^2 = \sum_{i = 1}^n \epsilon_i / (n - k - 1)$ is a estimator of $\sigma^2$, standard errors of the standard error of the regression are not directly provided.

However, the OLS estimator is also the same as the MLE estimator for $\beta$ (but not $\sigma$):
$$
\begin{aligned}[t]
p(y_1, \dots, y_n | \beta, \sigma, x_1, \dots, x_n) &= \prod_{i = 1}^n p(y_i | \beta, x_i) \\
&= \prod_{i = 1}^n N(y_i | x_i' \beta) \\
&= \prod_{i = 1}^n \frac{1}{\sigma \sqrt{2 \pi}} \left( \frac{-(y_i - x_i' \beta)}{2 \sigma^2} \right)
\end{aligned}
$$
so,
$$
\hat{\beta}_{MLE}, \hat{\sigma}_{MLE} = \arg\max_{b,s} \prod_{i = 1}^n N(y_i | x_i' b, s^2)  .
$$
And $\hat{\beta}_{MLE} = \hat{\beta}_{OLS}$.

Note that the OLS estimator is equivalent to the MLE estimator of $\beta$,
$$
\begin{aligned}[t]
\hat{\beta}_{MLE} &= \arg \max_{b} \prod_{i = 1}^n N(y_i | x_i' b, \sigma^2) \\
&=  \arg \max_{b} \prod_{i = 1}^n \frac{1}{\sigma \sqrt{2 \pi}} \exp \left( \frac{-(y_i - x_i' \beta)^2}{2 \sigma^2} \right) \\
&= \arg \max_{b} \log \left( \prod_{i = 1}^n \frac{1}{\sigma \sqrt{2 \pi}} \exp \left( \frac{-(y_i - x_i' \beta)}{2 \sigma^2} \right) \right) \\
&= \arg \max_{b} \sum_{i = 1}^n - \log \sigma - \frac{1}{2} \log 2 \pi + \frac{-(y_i - x_i' \beta)^2}{2 \sigma^2} \\
&= \arg \max_{b} \sum_{i = 1}^n  -(y_i - x_i' \beta)^2 \\
&= \arg \min_{b} \sum_{i = 1}^n  (y_i - x_i' \beta)^2  \\
&= \hat{\beta}_{OLS}
\end{aligned}
$$
However, the estimator of $\sigma^2_{MLE} \neq \sigma^2_{OLS}$.

### Bayesian Model with Improper priors

In Bayesian inference, our target is the posterior distribution of the parameters, $\beta$ and $\sigma$:  $p(\beta, \sigma^2 | y, X)$.

$$
p(\beta, \sigma | y, X) \propto p(y | \beta, \sigma) p(\beta, \sigma)
$$

For a Bayesian linear regression model, we'll need to specify distributions for $p(y | \beta, \sigma)$ and $p(\beta, \sigma)$.

**Likelihood:** $p(y_i | x_i, \beta, \sigma)$ suppose that the observations are distributed independent normal:
$$
y_i \sim \dnorm(\beta'x_i, \sigma^2)
$$

**Priors:** The model needs to specify a prior distribution for the parameters $(\beta, \sigma)$.
Rather than specify a single distribution for $\beta$ and $\sigma$, it will be easier to specify independent (separate) distributions for $\beta$ and $\sigma$.

We will use what are called an *improper uniform priors*.
An improper prior is,
$$
p(\theta) \propto C
$$
where $C$ is some constants.
This function puts an equal density on all values of the support of $\theta$.
This function is not a proper probability density function since $\int_{\theta \in \Theta} C d \theta = \infty$.
However, for some Bayesian models, the prior does not need to be a proper probability function for the posterior to be a probability function.
In this example we will put improper prior distributions on $\beta$ and $\sigma$.
$$
p(\beta, \sigma) = C
$$

$$
\begin{aligned}
p(\beta, \sigma | x, y) &\propto p(y| \beta, \sigma, x) p(\beta, \sigma, x) \\
&= \prod_{i = 1}^n N(y_i | x_i' \beta, \sigma^2) \cdot C \\
&\propto \prod_{i = 1}^n N(y_i | x_i' \beta, \sigma^2)
\end{aligned}
$$

Note that under the improper priors, the posterior is proportional to the likelihood,
$$
p(\beta, \sigma | x, y) \propto p(y | x, \beta, \sigma)
$$
Thus the MAP (maximum a posterior) estimator is the same as the MLE,
$$
\hat{\beta}_{MAP}, \hat{\sigma}_{MAP} = \arg\max_{\beta, \sigma} p(\beta, \sigma | x, y) = \arg \max_{\beta, \sigma} p(y | x, \beta, \sigma) = \hat{\beta}_{MLE}, \hat{\sigma}_{MLE}
$$

## Stan Model

Let's write and estimate our model in Stan.
Stan models are written in its own domain-specific language that focuses on declaring the statistical model (parameters, variables, distributions) while leaving the details of the sampling algorithm to Stan.

A Stan model consists of *blocks* which contain declarations of variables and/or statements.
Each block has a specific purpose in the model.

## Sampling Model with Stan

``` stan
functions {
    // OPTIONAL: user-defined functions
}
data {
    // read in data ...
}
transformed data {
    // Create new variables/auxiliary variables from the data
}
parameters {
    // Declare parameters that will be estimated
}
transformed parameters {
    // Create new variables/auxiliary variables from the parameters
}
model {
    // Declare your probability model: priors, hyperpriors & likelihood
}
generated quantities {
    // Declare any quantities other than simulated parameters to be generated
}
```

The file `lm0.stan` is a Stan model for the linear regression model previously defined.

```{r echo=FALSE, results="markup", comment="", lang = "stan"}
cat(readLines("stan/lm.stan"), sep = "\n")
```

```{r message=FALSE,results='hide'}
library("rstan")
mod1 <- stan_model("stan/lm.stan")
```

```{r cache=FALSE}
mod1
```

See the [Stan Modeling Language User's Guide and Reference Manual](http://mc-stan.org/documentation/) for details of the Stan Language.

**Note**Since a Stan model compiles to C++ code, you may receive some warning messages such as

```
/Library/Frameworks/R.framework/Versions/3.3/Resources/library/StanHeaders/include/stan/math/rev/core/set_zero_all_adjoints.hpp:14:17: warning: unused function 'set_zero_all_adjoints' [-Wunused-function]
    static void set_zero_all_adjoints() {
                ^
In file included from file1d4a4d50faa.cpp:8:
In file included from /Library/Frameworks/R.framework/Versions/3.3/Resources/library/StanHeaders/include/src/stan/model/model_header.hpp:4:
```

As long as your model compiles, you can ignore these compiler warnings (On the other hard, warnings that occur during sampling should not be ignored).
If the Stan model does not give you a syntax error when parsing the model, it should compile to valid C++.[^bugs][^c-warnings]
See

[bugs]: In the rare case that the Stan parser transpiles the Stan model to C++ but cannot compile the C++ code, it is a bug in Stan. Follow the [instructions](http://mc-stan.org/issues/) on how to inform the Stan developers about bugs.
[c-warnings]: The extended installation instructions for [MacOS/Linux](https://github.com/stan-dev/rstan/wiki/Installing-RStan-on-Mac-or-Linux) and [Windows](https://github.com/stan-dev/rstan/wiki/Installing-RStan-on-Windows) have instructions for adding compiler options to the R [Makevars](https://cran.r-project.org/doc/manuals/r-release/R-exts.html#Using-Makevars) file.

### Sampling

In order to sample from the model, we need to at least give it the values for the data to use: `n`, `k`, `y`, `X`, and the data associated with the priors.

```{r}
mod1_data <- list(
  y = Duncan$prestige,
  n = nrow(Duncan)
)
```

The data types in Stan are all numeric (either integers or reals), but they
include matrices and vectors. However, there is nothing like a data frame in
Stan. Whereas in the R function `lm` we can provide a formula and a data set
for where to look for objects, and the function will create the appropriate $X$
matrix for the regression, we will need to create that matrix
ourselves---expanding categorical variables to indicator variables, and
expanding interactions and other functions of the predictors. However, we need
to do that all manually. The function `r rdoc("stats", "model.matrix")` is the
workhorse function used in `lm` and many other R functions to convert a formula
into the matrix used in estimation.

```{r}
X <- model.matrix(prestige ~ type + income + education, data = Duncan)
mod1_data$X <- X
mod1_data$k <- ncol(X)
```

We still need to provide the values for the prior distributions.
For specific values of the prior distributions, assume uninformative priors for `beta` by setting the mean to zero and the variances to large numbers.
$$
\beta_k \sim \dnorm(0, 1000)
$$

```{r}
# mod1_data$b_loc <- 0
# mod1_data$b_scale <- 1000
```

For prior of the regression scale parameter $\sigma$, use a half-Cauchy distribution with a large scale parameter, which is a good choice for the priors of scale parameters.
<!--
In this case, `prestige` has values between 0 and 100.
This is like a proportion (actually, it is a proportion x 100), so ignoring the covariates, the maximum variance of a distribution would be if `prestige = 50`, when the standard deviation would be $\sqrt{p * (1 - p)} = 50$. So a scale parameter of 50 is appropriate,
-->
$$
\sigma \sim \dhalfcauchy(0, 50)
$$
```{r}
# mod1_data$sigma_scale <- 50
```

Now, sample from the posterior, using the function `sampling`:
```{r results='hide'}
mod1_fit <- sampling(mod1, data = mod1_data)
```

### Convergence Diagnostics and Model Fit

-   **Convergence Diagnostics:** Is this the posterior distribution that you
    were looking for? These don't directly say anything about how "good" the
    model is in terms representing the data, they are only evaluating how well
    the sampler is doing at sampling the posterior distribution of the given
    model. If there are problems with these, then the sample results do not
    represent the posterior distribution, and your inferences will be biased.

    -   `mcse`:
    -   `n_eff`:
    -   `Rhat`
    -   `divergences`

-   **Model fit:** Is this statistical model appropriate for the data?
    Or better than other models?

    -   Posterior predictive checks

    -   Information criteria:

        -   WAIC
        -   Leave-one-out Cross-Validation
