# Item Response Models

## Prerequisites {-}
```{r legislators_setup,message=FALSE}
library("pscl")
library("tidyverse")
library("forcats")
library("stringr")
library("rstan")
library("sn")
```

## Introduction

Item response theory (IRT) models are a measurment model for binary outcomes.
The data ($\Mat{y}$) are a $n \times m$ matrix.
The element $y_{ij} \in \{0, 1}$ is the response of subject $i \in 1, \dots, n$ answer to item $j \in 1, \dots, m$.
Common social science applications of this type of data include question outcomes for test-takers and yea-or-nay votes of legislators or judges.
The inferential problem is one of dimension reduction.

A common IRT model is the two-parameter model,
$$
\begin{aligned}[t]
y_{ij} &\sim \dbernoulli(\pi_{ij}) \\
\pi_{ij} &= \logit(\eta_{ij}) \\
\eta_{ij} &= \xi_i \beta_j - \alpha_j
\end{aligned}
$$
where

-   $\xi_i \in \R$ is a latent attribute of subject $i$.
    In test-taking, this would be the subject's ability.
    In voting, this would be the ideology or revealed preferences
    of the legislators.

-   $\alpha_j \in \R$ is called the *item difficulty*.
    Items with higher difficulty have a lower probability of $y_{ij}$.

-   $\beta_j \in \R$ is called the *item discrimination*.
    Items with higher magnitudes of discrimination are more
    highly related to the latent $\Vec{\xi}$ dimension.
    If $\beta_j = 0$, then the probability of $y_{ij}$ is 
    unrelated to the latent attribute of the subject.

Note that if $\Vec{xi}$ (or $\alpha_i$ and $\beta_j$) were known, then this model would be a typical binomial-logistic regression.
However, in this case, only the outcome $\Mat{y}$ is known.

Another way of writing this is as decomposing the $n \times m$ matrix $\Mat{\eta}$ into a $n \times 1$ vector of latent attributes $\Vec{\xi}$, a $m \times 1$ vector of item discriminations,
and a $m \times 1$ vector or item difficulties.
$$
\Mat{\eta} = \Vec{\xi} \Vec{\beta}' - \Vec{\alpha}
$$
The IRT model uses $2 m + n$ parameters to represent the $nm$ elements of $\Mat{\eta}$.
In many applied settings, $2 m + n \leq nm$; If $m \geq 1$, then this holds for all $n \geq 2m / (m - 1)$.


### Identification

The parameters $\xi$, $\beta$, and $\alpha$ are not identified in an IRT model.
This is common in many latent space models, including factor analysis.
In particular, there are three types of invariance:

1.  Scale invariance:
    $$
    \begin{aligned}[t]
    \mu_{i,j} &= \alpha_j + \beta_j \xi_i \\
             &= \alpha_j + \left(\frac{\beta_j}{c}\right) \left(\xi_i c \right) \\
             &= \alpha_j + \beta^*_j \xi^*_i
    \end{aligned}
    $$

1.  Location invariance:
    $$
    \begin{aligned}[t]
    \mu_{i,j} &= \alpha_j + \beta_j \xi_i \\
              &= \alpha_j - \beta_j c + \beta_j c + \beta_j \xi_i \\
             &= (\alpha_j - \beta_j c) + \beta_j (\xi_i + c) \\
             &= \alpha_j^{*} + \beta_j \xi^{*}_i
    \end{aligned}
    $$

1.  Rotation invariance:
    $$
    \begin{aligned}[t]
    \mu_{i,j} &= \alpha_j + \beta_j \xi_i \\
              &= \alpha_j  + \beta_j (-1) (-1) \xi_i \\
              &= \alpha_j  + (-\beta_j) (-\xi_i) \\
              &= \alpha_j + \beta_j^{*} \xi^{*}_i
    \end{aligned}
    $$

```{r legislators_identification}
xi <- c(-1, -0.5, 0.5, 1)
alpha <- c(1, 0, -1)
beta <- c(-0.5, 0, 0.5)
y <- matrix(c(1, 0, 1, 0, 0, 1, 1, 1, 1, 0, 1, 1), 3, 4)
k <- 1

list(sum(plogis(y - (alpha + beta %o% xi))),
     sum(plogis(y - (alpha + -beta %o% -xi))),
     sum(plogis(y - ((alpha - beta * k) + beta %o% (xi + k)))),
     sum(plogis(y - ((alpha + (beta / k) %o% (xi * k))))))
```

**Question:** For each of these: Which types of rotation does it solve?

1.  Fix one element of $\beta$.
1.  Fix one element of $\xi$.
1.  Fix one element of $\alpha$.
1.  Fix two elements of $\alpha$.
1.  Fix two elements of $\xi$.
1.  Fix two elements of $\beta$.

There are multiple ways to identify these models.

-   For $m = 1$, fix the values of two elements $\xi_{ij}$. 
-   For $m > 1$, 

# Legislators: Estimating Legislators' Ideal Points From Voting Histories {#legislators}

In political science, using this

## 109th Senate

This example models the voting of the [109th U.S. Senate](https://en.wikipedia.org/wiki/109th_United_States_Congress).
Votes for the 109th Senate (2015--2017) is included in the **pscl** package:

```{r}
data("s109", package = "pscl")
```

The `s109` object is not a data frame, so see its documentation for information about its structure.

This data includes all
[roll-call](https://en.wikipedia.org/wiki/Voting_methods_in_deliberative_assemblies)
votes, votes in which the responses of the senators are recorded.

For simplicity, the ideal point model uses binary responses, but the `s109`
data includes multiple [codes](http://voteview.com/senate109.htm) for response
to roll-calls.

| 0   | not a member                                              |
| 1   | Yea                                                       |
| 2   | Paired Yea                                                |
| 3   | Announced Yea                                             |
| 4   | Announced Nay                                             |
| 5   | Paired Nay                                                |
| 6   | Nay                                                       |
| 7   | Present (some Congresses, also not used some Congresses)  |
| 8   | Present (some Congresses, also not used some Congresses)  |
| 6   | Nay                                                       |
| 9   | Not Voting                                                |

In the data processing, we will aggregate the responses into "Yes", "No", and missing values.

-   `close`: Definition of non-lopsided votes in ; votes with between 35% and 65% yeas in which the parties are likely to whip members.
-   `lopsided`: Definition of lopsided votes used in W-NOMINATE and dropped. Votes with less than 2.5% or greater than 97.5% yeas.

```{r}
s109_vote_data <- as.data.frame(s109$vote.data) %>%
  mutate(rollcall = paste(session, number, sep = "-"),
         passed = result %in% c("Confirmed", "Agreed To", "Passed"),
         votestotal = yeatotal + naytotal,
         yea_pct = yeatotal / (yeatotal + naytotal),
         unanimous = yea_pct %in% c(0, 1),
         close = yea_pct < 0.35 | yea_pct > 0.65,
         lopsided = yea_pct < 0.025 | yea_pct > 0.975) %>%
  filter(!unanimous) %>%
  select(-unanimous) %>%
  mutate(.rollcall_id = row_number())

s109_legis_data <- as.data.frame(s109$legis.data) %>%
  rownames_to_column("legislator") %>%
  mutate(.legis_id = row_number(),
         party = fct_recode(party,
                            "Democratic" = "D",
                            "Republican" = "R",
                            "Independent" = "Indep"))

s109_votes <- s109$votes %>%
  as.data.frame() %>%
  rownames_to_column("legislator") %>%
  gather(rollcall, vote, -legislator) %>%
  # recode to Yea (TRUE), Nay (FALSE), or missing
  mutate(yea = NA,
         yea = if_else(vote %in% c(1, 2, 3), TRUE, yea),
         yea = if_else(vote %in% c(4, 5, 6), FALSE, yea)
         ) %>%
  filter(!is.na(yea)) %>%
  inner_join(dplyr::select(s109_vote_data, rollcall, .rollcall_id),
             by = "rollcall") %>%
  inner_join(dplyr::select(s109_legis_data, legislator, party, .legis_id), by = "legislator")

partyline <-
  s109_votes %>%
  group_by(.rollcall_id, party) %>%
  summarise(yea = mean(yea)) %>%
  spread(party, yea) %>%
  ungroup() %>%
  mutate(partyline = NA_character_,
         partyline = if_else(Republican < 0.1 & Democratic > 0.9,
                             "Democratic", partyline),
         partyline = if_else(Republican > 0.9 & Democratic < 0.1,
                             "Republican", partyline)) %>%
  rename(pct_yea_D = Democratic, pct_yea_R = Republican) %>%
  select(-Independent)

s109_vote_data <-
  left_join(s109_vote_data, partyline, by = ".rollcall_id")
```

## Identification by Fixing Legislator's Ideal Points

Identification of latent state models can be challenging. The first method for identifying ideal point models is to fix the values of two legislators.
These can be arbitrary, but if they are chosen along the ideological dimension of
interest it can help the substantive interpretation.

Since we \textit{a priori} know, or expect, that the primary ideological dimension is Liberal-Conservative [@PooleRosenthal2000a], I'll fix the ideal points of the two
party leaders in that congress.
In the 109th Congress, the Republican party was the majority party and [Bill Frist](https://en.wikipedia.org/wiki/Bill_Frist) (Tennessee) was the majority (Republican) leader, and [Harry Reid](https://en.wikipedia.org/wiki/Harry_Reid) (Nevada) wad the minority (Democratic) leader:
$$
\begin{aligned}[t]
\xi_\text{FRIST (R TN)} & = 1 \\
\xi_\text{REID (D NV)} & = -1
\end{aligned}
$$

For all of those give a weakly informative prior to the ideal points, and item difficulty and discrimination parameters,
$$
\begin{aligned}[t]
\xi &\sim \mathsf{Normal}(0., 10) \\
\alpha_{j} &\sim \mathsf{Normal}(0, 10) \\
\beta_{j} &\sim \mathsf{Normal}(0, 2.5) && j \in 1, \dots, J
\end{aligned}
$$

```{r legislators_mod_ideal_point_1,results='hide',cache.extra=tools::md5sum("stan/ideal_point_1.stan")}
mod_ideal_point <- stan_model("stan/ideal_point.stan")
```

```{r echo=FALSE,results,comment=''}
cat(readLines("stan/ideal_point.stan"), sep = "\n")
```

Create a data frame with the fixed values for identification.

```{r legislators_xi_init_1}
ideal_points <-
  s109_legis_data %>%
  mutate(xi = if_else(legislator == "FRIST (R TN)", 1,
                      if_else(legislator == "REID (D NV)", -1, NA_real_)),
         init = if_else(party == "Republican", 1,
                        if_else(party == "Democratic", -1, 0)))         
```

Setup the data for Stan.
```{r legislators_data_1}
legislators_data <-
  dplyr::lst(y = as.integer(s109_votes$yea),
    P = length(y),
    y_idx_leg = as.integer(s109_votes$.legis_id),
    y_idx_vote =  as.integer(s109_votes$.rollcall_id),
    y_obs = length(y),
    N = max(s109_votes$.legis_id),
    M = max(s109_votes$.rollcall_id),
    # priors
    loc_alpha = 0,
    scale_alpha = 10,
    loc_beta = 0,
    scale_beta = 2.5,
    N_obs = sum(!is.na(ideal_points$xi)),
    idx_xi_obs = which(!is.na(ideal_points$xi)),
    xi_obs = ideal_points$xi[!is.na(ideal_points$xi)],
    idx_xi_param = which(is.na(ideal_points$xi)),
    loc_xi = 5,
    scale_xi = 20,
    df_xi = 4
  )
```

```{r legislators_init_1}
legislators_init <- list(
  list(xi_param = ideal_points$init[is.na(ideal_points$xi)])
)
```

```{r legislators_fit,results='hide'}
legislators_fit <- sampling(mod_ideal_point, data = legislators_data,
                            chains = 1, iter = 500, refresh = 100)
```

Extract the ideal point data:
```{r legislator_summary}
legislator_summary <-
  bind_cols(s109_legis_data,
           as_tibble(summary(legislators_fit, par = "xi")$summary)) %>%
  mutate(legislator = fct_reorder(legislator, mean))
```

```{r legislator_plot_1,fig.height=8,fig.width=4,fig.cap="Estimated Ideal Points of the Senators of the 109th Congress"}
ggplot(legislator_summary,
       aes(x = legislator, y = mean,
           ymin = `2.5%`, ymax = `97.5%`, colour = party)) +
  geom_pointrange() +
  coord_flip() +
  scale_color_manual(values = c(Democratic = "blue", Independent = "gray",
                                Republican = "red")) +
  labs(y = expression(xi[i]), x = "", colour = "Party") +
  theme(legend.position = "bottom")
```

## References

See the links in [Education Research Using Stan](https://education-stan.github.io/)
for examples of IRT models with Stan.
In particular, the package **edstan** and **idealstan** have education IRT and ideal point  models implemented in Stan.

There are numerous other IRT models implementations in R. 
In the **MCMCpack** package see functions `MCMCirt1d`, `MCMCirtKd`, `MCMCirtKdRob`, and
`MCMCirtHir1d`.

[^legislators-src]: Example derived from Simon Jackman, "Legislators: estimating legislators' ideal points from voting histories (roll call data)", *BUGS Examples,* 2007-07-24, [URL](https://web-beta.archive.org/web/20070724034141/http://jackman.stanford.edu:80/mcmc/legislators.odc).
