# Multilevel Models

```{r message=FALSE}
library("rstan")
library("rstanarm")
library("tidyverse")
```

Multilevel models are commonly used hierarchical model.
They extend (generalized) linear models to include coefficients that vary by discrete groups.

Suppose that there are $i = 1, dots, n$ observations, and each observation is in one of $j = 1, \dots, J$ groups.
Let $j[i]$ be the group for
$$
\begin{aligned}[t]
y_i &\sim \dnorm(\alpha_{j[i]} + \beta_{j[i]} x_i, \sigma^2) \\
  \begin{bmatrix}
  \alpha_j \\
  \beta_j
  \end{bmatrix}
& \sim
\dnorm
\left(
  \begin{bmatrix}
  \mu_\alpha \\
  \mu_\beta
  \end{bmatrix},
\Omega
\right)
\end{aligned} .
$$

*Pooled model*: All coefficients are common between groups. This is equivalent to a linear model.
$$
\begin{aligned}[t]
y_i &\sim \dnorm(\alpha + \beta x_i, \sigma^2) \\
\begin{bmatrix}
\alpha \\
\beta
\end{bmatrix}
&\sim
\dnorm
\left(
  \begin{bmatrix}
  \mu_{\alpha} \\
  \mu_{\beta}
  \end{bmatrix},
  \Omega
\right)
\end{aligned}
$$

*Pooled model*: All coefficients are common between groups. This is equivalent to a linear model.
$$
\begin{aligned}[t]
y_i &\sim \dnorm(\alpha + \beta x_i, \sigma^2) \\
\end{aligned}
$$
*Varying-intercept*: The slope coefficients ($\beta$) are common between groups, but the intercepts ($\alpha_j$) vary by group.
$$
\begin{aligned}[t]
y_i &\sim \dnorm(\alpha_{j[i]} + \beta x_i, \sigma^2) \\
\end{aligned}
$$
*Varying-slope model*: The groups share a common intercept, $\alpha$, but the slope coefficient ($\beta$), varies between groups. This is less common since it is hard to think of cases when it is appropriate.
$$
\begin{aligned}[t]
y_i &\sim \dnorm(\alpha + \beta_{j[i]} x_i, \sigma^2) \\
\end{aligned}
$$
These models go by different names in different literatures:
*hierarchical (generalized) linear models*, *nested data models*,
*mixed models*, *random coefficients*, *random-effects*,
*random parameter models*,  *split-plot designs* [^mlm-names].

The model can be extended to other cases:

-   generalized linear models
-   multiple parameters

One of the difficulties in these models is the prior to the covariance matrix, $\Omega$.

```{r echo=FALSE, fig.cap="Visual representation of hierarchical models"}
n <- 20
b <- 1
a <- 0
Sigma <- matrix(c(1, 0, 0, 1), nrow = 2) * 0.5
slopes <- mvtnorm::rmvnorm(n, c(a, b), Sigma)  %>%
  as_tibble() %>%
  set_names(c("intercept", "slope"))

bind_rows(
  mutate(slopes, x = -2,
         xend = 2,
         y = a + b * x,
         yend = a + b * xend,
         type = "No-pooling"),
  mutate(slopes, x = -2,
         xend = 2,
         y = intercept + b * x,
         yend = intercept + b * xend,
         type = "Varying intercept"),
  mutate(slopes, x = -2,
         xend = 2,
         y = a + slope * x,
         yend = a + slope * xend,
         type = "Varying Slopes"),
  mutate(slopes,
         x = -2,
         xend = 2,
         y = intercept + slope * x,
         yend = intercept + slope * xend,
         type = "Varying slope and intercept")
) %>%
  ggplot(aes(x = x, xend = xend, y = y, yend = yend)) +
  geom_segment() +
  facet_wrap(~ type, ncol = 2) +
  theme_void()

```

## Example: Radon

This example models the presence of radon in houses in Minnesota which appears in @GelmanHill2007a and @BDA3.
This is partly derived from a [Stan Case Study](http://mc-stan.org/documentation/case-studies/radon.html), which uses `PyStan` instead of **rstan**.

### Data

The `r rdoc("radon", "rstanarm")` data is included in the `r rpkg("rstanarm")` package.
```{r}
data("radon", package = "rstanarm")
glimpse(radon)
```

The data consist of `r nrow(radon)`  observations of radon levels of houses from `r nlevels(radon$county)` counties.

```{r radon_county}
radon_county <- radon %>%
  group_by(county) %>%
  summarise(log_radon_mean = mean(log_radon),
            log_radon_sd = sd(log_radon),
            log_uranium = mean(log_uranium),
            n = length(county))

```

```{r fig.asp=2}
ggplot() +
  geom_point(data = radon,
             mapping = aes(y = log_radon,
                           x = fct_reorder(county, log_radon, mean))) +
  geom_point(data = radon_county,
             mapping = aes(x = fct_reorder(county, log_radon_mean),
                           y = log_radon_mean),
             colour = "red") +
  coord_flip() +
  labs(y = "log(radon)", x = "")
```

Relationship between mean and sample size
```{r}
ggplot(radon_county, aes(y = log_radon_mean, x = log2(n))) +
  geom_point()
```

### Varying Intercepts Models

Consider the general model with an intercept for each county representing the baseline average of the county:
$$
\begin{aligned}
y_i &\sim  N(\mu_i, \sigma^2) \\
\mu_i &= \alpha_{j[i]} + \beta x_i
\end{aligned}
$$
where $j[i]$ means that observation $i$ is in county $j \in (1, \dots, `r nlevels(radon$county)`)$.

In this particular example, $y = \mathtt{log_radon}$ and $x = \mathtt{basement}$.
$$
\begin{aligned}
\mathtt{log\_radon}_i &\sim  N(\mu_i, \sigma^2) \\
\mu_i &= \alpha_{j[i]} + \beta~\mathtt{basement}_i
\end{aligned}
$$

We can put a prior distribution on $\alpha_{j[i]}$,
$$
\begin{aligned}[t]
\alpha_{j} &\sim N(\gamma, \tau) & \text{for $i \in (1, \dots, `r nlevels(radon$county)`)}
\end{aligned}
$$
This parameterization nests common cases,

*Complete pooling:* When $\tau \to 0$, the intercepts are the same,
$$
\begin{aligned}[t]
\alpha_j &= \gamma  & \text{for all $j$.}
\end{aligned}
$$

*No pooling:* When $\tau \to \infty$, prior distribution on the intercepts is equivalent to an improper normal distribution, and there is no shrinkage,
$$
p(\alpha_j) \propto 1,
$$
for all $j$.

*Partial pooling:* When $\tau$ is a parameter, the amount of shrinkage can be estimated from the data.

### Varying Intercept Model

### Varying Slope Model

$$
\begin{aligned}
\mathtt{log\_radon}_i &\sim  N(\mu_i, \sigma^2) \\
\mu_i &= \alpha_{j[i]} + \beta_{j[i]}~\mathtt{basement}_i
\end{aligned}
$$

### Group Level Predictors

The `radon` dataset also contains the county-level measurements of `uranium`.

One way to include county level measurements is to model the county-level intercepts.
The values of each county intercept is a function of the county-level uranium.
$$
\begin{aligned}
\mathtt{log\_radon}_i &\sim  N(\mu_i, \sigma^2) \\
\mu_i &= \alpha_{j[i]} + \beta_{j[i]}~\mathtt{basement}_i
\alpha_{j} \sim  N(\gamma_0 + \gamma_1~\mathtt{log\_uranium}_j, \tau)
\end{aligned}
$$

Alternatively, we can model model the county-level intercepts.
The values of each county intercept is a function of the county-level uranium.
$$
\begin{aligned}
\mathtt{log\_radon}_i &\sim  N(\mu_i, \sigma^2) \\
\mu_i &= \alpha_{j[i]} + \beta_{j[i]}~\mathtt{basement}_i \\
\alpha_{j} &\sim  N(\gamma_0 + \gamma_1~\mathtt{log\_uranium}_j, \tau)
\end{aligned}
$$

### lme4

In R, the most widely used package to estimate mixed-effects models is **lme4**.
This estimates models using maximum likelihood or restricted maximum likelihood methods (REML).
This will be faster than using full-Bayesian methods but also underestimate the uncertainty, as well as being a worse approximation of the posterior.
Additionally, in frequentist inference, the meaning of the random effects is different; they are nuisance parameters and not given standard errors.

See @Bates2010a and @BatesMaechlerBolkerEtAl2014a for introductions to mixed-effects models with **lme4**.
These are also good introductions to classical approaches to mixed effects models.

```{r message=FALSE}
library("lme4")
```

Complete pooling
```{r}
fit_pooled <- lm(log_radon ~ county + floor, data = radon)
```
County-varying intercepts with no-pooling
```{r}
fit_intercept_nopool <- lm(log_radon ~ floor, data = radon)
```
County-varying intercepts with partial-pooling
```{r}
fit_intercept_partial <- lmer(log_radon ~ (1 | county) + floor, data = radon)
```
Varying slopes with no pooling:
```{r}
fit_slope_nopool <- lm(log_radon ~ county * floor, data = radon)
```
Varying slopes with partial pooling:
```{r}
fit_slope_partial <- lmer(log_radon ~ (1 + floor | county), data = radon)
```

Including a county-level variable (`log_uranium`) in various models:

With no-pooling,
```{r}
fit_slope_partial <- lm(log_radon ~ floor + log_uranium, data = radon)
```
With varying-intercepts
```{r}
fit_slope_partial <- lmer(log_radon ~ (1 | county) + floor + log_uranium,
                          data = radon)
```
With varying-intercepts and slopes,
```{r}
fit_slope_partial <- lmer(log_radon ~ (1 + floor | county) +  log_uranium,
                          data = radon)
```

### rstanarm

Some multilevel models can also be estimated using the **rstanarm** functions `stan_glmer` and `stan_lmer`.
These functions have syntax similar to **lme4** functions, but estimate the mixed models using Bayesian methods with Stan.

Complete pooling
```{r}
fit_pooled <- stan_glm(log_radon ~ county + floor, data = radon)
```
County-varying intercepts with no-pooling
```{r}
fit_intercept_nopool <- stan_glm(log_radon ~ floor, data = radon)
```
County-varying intercepts with partial-pooling
```{r}
fit_intercept_partial <- stan_glmer(log_radon ~ (1 | county) + floor,
                                    data = radon)
```
Varying slopes with no pooling. *There is an error estimating this*
```{r eval=FALSE}
fit_slope_nopool <- stan_glm(log_radon ~ -1 + county + county:floor,
                             data = radon,
                             prior = normal(scale = 1))
```
Varying slopes with partial pooling:
```{r}
fit_slope_partial <- stan_glmer(log_radon ~ (1 + floor | county), data = radon)
```

Including a county-level variable (`log_uranium`) in various models:

With no-pooling,
```{r}
fit_slope_partial <- stan_glm(log_radon ~ floor + log_uranium, data = radon)
```
With varying-intercepts
```{r}
fit_slope_partial <-
  stan_glmer(log_radon ~ (1 | county) + floor + log_uranium,
             data = radon)
```
With varying-intercepts and slopes,
```{r}
fit_slope_partial <-
  stan_glmer(log_radon ~ (1 + floor | county) +  log_uranium,
             data = radon)
```

## Pooling of Hierarchical Parameters

This is easiest understood in the case of a model of group means,
$$
\begin{aligned}[t]
y &\sim \dnorm(\mu_{j[i]}, \sigma^2) \\
\mu_{j} &\sim \dnorm(\gamma, \tau^2) .
\end{aligned}
$$
Each group has size $n_j$.

| Sample size, $n_j$              | Estimate of $\hat{\mu}_j$                                       |
| ------------------------------- | --------------------------------------------------------------- |
| $n_j = 0$                       | $\hat{\mu}_j = \gamma$ (complete pooling)                       |
| $n_j < \frac{\sigma^2}{\tau^2}$ | $\hat{\mu}_j$ closer to $\gamma$                                |
| $n_j = \frac{\sigma^2}{\tau^2}$ | $\hat{\mu}_j = \frac{1}{2} \bar{y}_j + \frac{1}{2} \gamma$      |
| $n_j > \frac{\sigma^2}{\tau^2}$ | $\hat{\mu}_j$ closer to $\bar{y}_j$                             |
| $n_j = \infty$                  | $\hat{\mu}_j = \bar{y}_j$ (no pooling)                          |

If the hyperparameters were known, the posterior of $\mu_j$ is
$$
\mu_j | y, \gamma, \sigma, \tau \sim \dnorm(\hat{\mu}_j, V_j)
$$
where
$$
\begin{aligned}[t]
\hat{\mu}_j &= \frac{\frac{n_j}{\sigma^2} \bar{y}_j + \frac{1}{\tau^2} \gamma}{\frac{n_j}{\sigma^2} + \frac{1}{\tau^2}} \\
V_j &= \frac{1}{\frac{n_j}{\sigma^2} + \frac{1}{\tau^2}}
\end{aligned}
$$

Some crude estimates given $\mu_j$.

The *data variance*, $\sigma^2$, is the residual variance,
$$
\E(\sigma^2 | y, \mu)  = \frac{1}{n} \sum_{i = 1}^n (y - \mu_{j[i]})^2 .
$$
The global mean is approximately the average of the group-level means,
$$
\begin{aligned}[t]
\E(\gamma | y, \mu) &= \frac{1}{J} \sum_{i = 1}^n \mu_j \\
\Var(\gamma | y, \mu) &= \frac{1}{J} \tau^2
\end{aligned}
$$
The group level variance is $\tau^2$ is,
$$
\E(\tau^ | y, \mu) = \frac{1}{J} \sum_{j = 1}^J (\mu_j - \gamma)^2
$$

## ANOVA

**TODO**

## Time-Series Cross Section

**TODO**

A common application for these models are Time-Series Cross-Section (TSCS) or panel models.
In this case, both the time and units can be modeled.

 ## Extensions

-   Including group-level covariates

-   Prior distributions

-   Prediction

    -   new obs in existing groups
    -   new group
    -   new obs in new group

-   Modeling correlation between intercept and slopes

-   Non-nested models

## Miscellaneous

### How many groups?

In classical discussions of multi-level or hierarchical models,
a common question is how many groups are required to be able to use random effects vs. fixed effects.

As noted earlier, random effects estimates the variance between group means. If there are few groups, there is not much information available to estimate this variance.
As such, random effects is not much different than fixed effects.

This literature provides many different rules of thumb for the number of groups necessary to be able to use random effects: 8, 10, 30, 50, or 100 [@Stegmueller2013a, p. 749].

@Stegmueller2013a finds that Bayesian method produces better multi-level-models than maximum likelihood methods for all numbers of groups.
ML methods do not suffer severe bias above 10-15 groups.
Bayesian point estimates are biased for smaller numbers of groups, but less than the ML.
Additionally, the Bayesian methods have better frequentist coverage than ML methods.

@BeckKatz2007a show that ML random coefficient models are superior in terms of efficiency to many types of pooled and un-pooled estimators in small samples.

### Correlation between Predictors and Errors

@BafumiGelman2006a analyze this case.

The standard suggestion in frequentist literature is to use a Hausman test
where the null hypothesis is that random effects are consistent. However,
@ClarkLinzer2014a note that in small samples this is likely to fail to reject
random effects; and in large samples, random effects behave like fixed effects
anyways.

## References

Texts and chapters on multi-level analysis:

-   Bayesian

    -   @GelmanHill2007a [Ch. 11-17].
    -   @BDA3 [Ch 5] "Hierarchical Models"
    -   @BDA3 [Ch 15] "Hierarchical Linear Models"
    -   @Jackman2009a [CHh. 7]
    -   @Draper2008a

-   Frequentist

    -   @Goldstein2011a
    -   @SnijdersBosker2011a
    -   @Rabe-HeskethSkrondal2012a
    -   @Jiang2007a

Stan model examples:

-   Stan models for [ARM](https://github.com/stan-dev/example-models/wiki/ARM-Models)
-   <http://mc-stan.org/documentation/case-studies/radon.html>
-   <https://biologyforfun.wordpress.com/2016/12/08/crossed-and-nested-hierarchical-models-with-stan-and-r/>

Examples of multilevel models

-   @Western1998a: economic growth for OECD countries

-   @GelmanKing1993a: US election polling

-   @ParkGelmanBafumi2004a: multi-level models of opinion polls combined with post-stratification to extrapolate national opinion surveys to regions.

-   @SteenbergenJones2002a: mostly an intro/review of MLM, but uses the
    cross-country Eurobarometer to model support for the EU

-   @GelmanShorBafumiEtAl2007a: state-level opinion polls

-   @RaudenbushBryk2001a: student performance with student and school-level indicators

-   @Gilardi2010a: policy diffusion

-   @ORourkeSinnott2006a: attitudes toward immigration

-   @AndersenFetner2008a: ethnic and social tolerance

-   @Weldon2006a: ethnic and social tolerance

-   @Arzheimer2009a: right-wing voting

-   @HoogheReeskensStolleEtAl2009a: social and political trust

-   @AndersonSinger2008a: satisfaction with democracy

-   @MeerDethScheepers2009a: political participation

-   @IversenRosenbluth2006a: political economy of the gender wage gap

-   @HoogheMarks2004a: support for European integration

-   @LaxPhillips2009a: American politics using states and neighborhoods

-   @Voeten2008a: judicial decision making

-   @FranchinoHoeyland2009a: legislative politics

-   @DenisovaEllerFryeEtAl2009a: politics of economic reforms

-   @AitkinLongford1986a, @GoldsteinYangOmarEtAl2000a,
    @GoldsteinRasbashYangEtAl1993a: education

-   @GoldsteinYangOmarEtAl2000a: medicine

[^mlm-names]: <https://en.wikipedia.org/wiki/Multilevel_model>
