# Posterior Inference

## Prerequisites

The `r rpkg("haven")` package is used to read Stata `.dta` files.
```{r}
library("rubbish")
library("haven")
```

## Introduction

The posterior distribution is the probability distribution $\Pr(\theta | y)$.

One we have the posterior distribution, or more often a sample from the posterior distribution, it is relatively easy to perform inference on any function of the posterior.

Common statistics used to summarize the posterior distribution:

- mean: $\E(p(\theta | y)) \approx \frac{1}{S} \sum_{i = 1}^S \theta^{(s)}$
- median: $\median(p(\theta | y)) \approx \median \theta^{(s)}$
- quantiles: 2.5%, 5%, 25%, 50%, 75%, 95%, 97.5%
- credible interval:

    - central credible interval: the interval between the p/2% and 1 - p/2% quantiles
    - highest posterior density interval: the narrowest interval containing p% of distribution

- marginal densities

## Functions of the Posterior Distribution

It is also easy to conduct inference on functions of the posterior distribution.

Suppose $\theta^{(1)}, \dots, \theta^{(S)}$ are a sample from $p(\theta | y)$, the
$f(\theta^{(1)}), \dots, f(\theta^{(S)})$ are a sample from $p(f(\theta) | y)$.

This is not easy for methods like MLE that produce point estimates:

- Even in OLS, non-linear functions coefficients generally require either the Delta method or bootstrapping to calculate confidence intervals.
- @BerryGolderMilton2012a, @Goldera,@BramborClarkGolder2006a discuss calculating confidence intervals
- See @Rainey2016b on "transformation induced bias"
- See @Carpenter2016a on how reparameterization affects point estimates; this is a Stan Case study with working code

## Marginal Effects

### Example: Marginal Effect Plot for X

@BerryGolderMilton2012a replicates @Alexseev2006a as an example of a model with an interaction between $X$ and $Z$.
$$
Y = \beta_0 + \beta_x X + \beta_z Z + \beta_{xz} X Z + \epsilon
$$
In this case, the hypothesis of interest involves the marginal effect of $X$ on $Y$,
$$
\frac{\partial \E(Y|.)}{\partial X} = \beta_z + \beta_{xz} Z
$$
Since there is an interaction, the marginal effect of $X$ is not simply
the coefficient $\beta_z$, but is a function of another predictor, $Z$.
Point estimates of the marginal effects with interactions are relatively easy to construct, but confidence intervals for the MLE estimates quickly involve multiple terms and either the Delta method approximation or bootstrapping to calculate.

We will consider this problem from a Bayesian estimation perspective, and calculate point estimates (posterior mean) and credible intervals of the marginal effects.

The particular example is @Alexseev2006a, which analyzes how changes in the ethnic composition of Russian regions affected the vote share of the extreme Russian nationalist Zhirinovsky Bloc in 2003 Russian State Duma elections.[^alexseev1]
```{r}
alexseev <- read_dta("data/alexseev.dta")
```

[alexseev1]: Some of the replication code and material can be found on [Matt Golder's website](http://mattgolder.com/interactions).

One claim of @Alexseev2006a was that support for anti-immigrant parties depends on the percentage of the population of the dominant ethnic group (Slavic) and the change in the percentage the non-dominant share.
To test that hypothesis, @Alexseev2006a estimates the following model,
$$
\begin{multline}
\mathtt{xenovote}_i = \beta_0 + \beta_s \mathtt{slavicshare}_i +
\beta_{n} \mathtt{changenonslav} + \\
\beta_{sn} (\mathtt{slavicshare}_i \times \mathtt{changenonslav}_i) +
\gamma z_{i} + \epsilon_{i},
\end{multline}
$$
where $z_i$ is a vector of control variables.

-   `xenovote`: Xenophobic voting. Share of vote for the [Zhirinovsky Bloc](https://en.wikipedia.org/wiki/Vladimir_Zhirinovsky).
-   `slavicshare`: *Slavic Share.* Proportion Slavic in the district.
-   `changenonslav`: $\Delta$ *non-Slavic Share* Change in the proportion of non-Slavic groups in the region.

The model was estimated by OLS,[^alexseev2]
```{r}
mod_f <- (xenovote ~ slavicshare * changenonslav + inc9903 + eduhi02 + unemp02 +
            apt9200 + vsall03 + brdcont)
lm(mod_f, data = alexseev)
```

[^alexseev2]: The original model used cluster robust standard errors, which will be ignored for now.

Use the `lm_preprocess` function in the [rubbish](https://jrnold.github.com/rubbish) package to turn the model formula into a list with relevant data.
```{r}
mod_data <- lm_preprocess(mod_f, data = alexseev)[c("X", "y")]

mod_data <- within(mod_data, {
  n <- nrow(X)
  k <- ncol(X)
  M <- 100
  changenonslav <- seq(min(X[, "changenonslav"]),
                       max(X[, "changenonslav"]),
                       length.out = M)
  idx_b_slavicshare <- which(colnames(X) == "slavicshare")
  idx_b_slavicshare_changenonslav <-  # nolint
    which(colnames(X) == "slavicshare:changenonslav")
  b_loc <- 0
  # data appropriate prior
  b_scale <- max(apply(X, 2, sd)) * 3
  sigma_scale <- sd(y)
})
```
```{r include = FALSE}
mod <- stan_model("stan/ex-alexseev.stan")
```
```{r cache=FALSE}
mod
```

```{r include=FALSE}
mod_fit <- sampling(mod, data = mod_data)
mod_fit
```


The function `r rdoc("rstan", "extract")` extracts parameters from
the `r rdoc("rstan", "stanfit")` object as a list with an element for each parameter.
```{r}
dydx_all <-
  rstan::extract(mod_fit, pars = "dydx")$dydx
```
To make it easier to use with `ggplot`, convert it to a data frame with columns `.id` (number), `changeonslav` (original value of `changeonslav` passed to the sampler), and `value` (the value of the marginal effect).
```{r}
dydx_all <-
  dydx_all %>%
  as.tibble() %>%
  mutate(.iter = row_number()) %>%
  gather(.id, value, -.iter) %>%
  # merge with original values of changenonslav
  left_join(tibble(.id = paste0("V", seq_along(mod_data$changenonslav)),
            changenonslav = mod_data$changenonslav),
            by = ".id")
```

Since values for a marginal effect line is generated for each iteration, the posterior distribution of $\partial E(\mathtt{xenovote}|.) / \partial slavicshare$ can be plotted as lines.
Since the number of lines would be too many to display effectively, plot `r 2 ^ 8` of them:
```{r}
dydx_all %>%
  filter(.iter %in% sample(unique(.iter), 2 ^ 8)) %>%
  ggplot(aes(x = changenonslav, y = value, group = .iter)) +
  geom_line(alpha = 0.3) +
  ylab("Marginal effect of slavic share") +
  xlab(paste(expression(Delta, "non-Slavic Share")))
```

Alternatively, we can summarize the posterior distribution of the marginal effects with a line (posterior mean) and credible interval regions (50%, 90%):
```{r}
dydx_summary <-
  dydx_all %>%
  group_by(changenonslav) %>%
  summarise(mean = mean(value),
            q5 = quantile(value, 0.05),
            q25 = quantile(value, 0.25),
            q75 = quantile(value, 0.75),
            q95 = quantile(value, 0.95))
ggplot() +
  modelr::geom_ref_line(h = 0) +
  geom_ribbon(data = dydx_summary,
              mapping = aes(x = changenonslav, ymin = q5, ymax = q95),
              alpha = 0.2) +
  geom_ribbon(data = dydx_summary,
              mapping = aes(x = changenonslav, ymin = q25, ymax = q75),
              alpha = 0.2) +
  geom_line(data = dydx_summary,
            mapping = aes(x = changenonslav, y = mean),
            colour = "blue") +
  geom_rug(data = alexseev, mapping = aes(x = changenonslav), sides = "b") +
  ylab("Marginal effect of slavic share") +
  xlab(expression(paste(Delta, "non-Slavic Share")))
```
The plot above also includes a rug with the observed values of `changenonslav` in the sample.

-  **Q:** For each value of `changenonslav`, what is the probability that the marginal effect of `slavicshare` is greater than 0?
-  **Q:** Re-estimate the model, but calculate the marginal effect of `slavicshare` for all observed values of `changenonslav` in the sample. For each observation, calculate the probability that the marginal effect is greater than 0. What proportion of observations is the probability that the marginal effect is greater than zero.
-  **Q:** Suppose you want to calculate the expected probability that the marginal effect of `slavicshare` is greater than zero in the sample. Let $\theta^{S}_i$ be the parameter for the marginal effect of `slavicshare` on the `xenovote`. Consider these two calculations:
    $$
    \frac{1}{N} \sum_{i = 1}^n \left( \frac{1}{S} \sum_{s = 1}^S I(\theta^{(s)}_i > 0) \right)
    $$
    and
    $$
    \frac{1}{S} \sum_{s = 1}^S \left( \frac{1}{N} \sum_{i = 1}^N I(\theta^{(s)}_i > 0) \right) .
    $$
    Are they the same? What are their substantive interpretations?
-   **Q:** Construct the same plot but for Figure 5(b) in @BerryGolderMilton2012a, which displays the marginal effects of $\Delta$ *non-Slavic* on *Xenophobic* voting.
