# Heteroskedasticity and Robust Regression


`r rpkg("VGAM")` is needed for the Laplace distribution.
```{r message=FALSE}
library("rubbish")
library("VGAM")
```


## Linear Regression with Student t distributed errors


Like OLS, Bayesian linear regression with normally distributed errors is sensitive to outliers.
The normal distribution has narrow tail probabilities.


This plots the normal, Double Exponential (Laplace), and Student-t (df = 4) distributions all with mean 0 and scale 1, and the surprise ($- log(p)$) at each point.
Higher surprise is a lower log-likelihood.
Both the Student-t and Double Exponential distributions have surprise values well below the normal in the ranges (-6, 6).[^tailareas]
This means that outliers impose less of a penalty on the log-posterio models using these distributions, and the regression line would need to move less to incorporate those observations since the error distribution will not consider them as unusual.

[^tailareas]: The Double Exponential distribution still has a thinner tail than the Student-t at higher values.

```{r}
z <- seq(-6, 6, length.out = 100)
bind_rows(
  tibble(z = z,
         p = dnorm(z, 0, 1),
         distr = "Normal"),
  tibble(z = z,
         p = dt(z, 4),
         distr = "Student-t (df = 4)"),
  tibble(z = z,
         p = VGAM::dlaplace(z, 0, 1),
         distr = "Double Exponential")) %>%
  mutate(`-log(p)` = -log(p)) %>%
  ggplot(aes(x = z, y = `-log(p)`, colour = distr)) + 
  geom_line()
       

```


```{r include=FALSe}
mod_t <- stan_model("stan/rlm.stan")
```
```{r}
unionization <- read_tsv("data/western1995/unionization.tsv",
         col_types = cols(
              country = col_character(),
              union_density = col_double(),
              left_government = col_double(),
              labor_force_size = col_number(),
              econ_conc = col_double()
            ))
mod_t_data <- preprocess_lm(union_density ~ left_government + log(labor_force_size) + econ_conc, data = unionization)
                                   
mod_data <- within(mod_data, {
  b_loc <- 0
  b_scale <- 100
  sigma_scale <- sd(y)
})

```

```{r}
mod_data$X
```


The `max_treedepth` parameter needed to be increased because in some runs it was hitting the maximum treedepth.
This is likely due to the wide tails of the Student t distribution.
```{r}
mod_t_fit <- sampling(mod_t, data = mod_data, control = list(max_treedepth = 11))
```

```{r output=FALSE}
summary(mod_t_fit, pars = c("nu", "sigma", "b"))$summary
```

Compare those results when using a model with 
```{r include=FALSE}
mod_normal <- stan_model("stan/lm.stan")
mod_normal_fit <- sampling(mod_normal, data = mod_data)
```

```{r}
summary(mod_normal_fit, pars = c("b", "sigma"))$summary
```

Alternatively, the Double Exponential (Laplace) distribution can be used for the errors.
This is the equivalent to least quantile regression, where the regression line is the median (50% quantile)
```{r}
mod_dbl_exp <- stan_model("stan/lms.stan")
mod_dbl_exp_fit <- sampling(mod_dbl_exp, data = mod_data)
```
```{r}
summary(mod_dbl_exp_fit, par = c("b", "sigma"))$summary
```



## Heteroskedasticity

In applied regression, heteroskedasticity consistent (HC) or robust standard errors are often used.

However, there is straighforwardly direct translation of HC standard error to regression model this in a Bayesian setting. The sandwich method of estimating HC errors uses the same point estimates for the regression coefficients as OLS, but estimates the standard errors of those coefficients in a second stage from the OLS residuals. 
Disregarding differences in frequentist vs. Bayesian inference, it is clear that a direct translation of that method could not be fully Bayesian since the coefficients and errors are not estimated jointly.

In a linear normal regression model with heteroskedasticity, each observation has its own scale parameter, $\sigma_i$,
$$
\begin{aligned}[t]
y_i &\sim \dnorm(X \beta, \sigma_i) .
\end{aligned}
$$
It should be clear that without proper priors this model is not identified, meaning that the posterior distribution is improper.
To estimate this model we have to apply some model to the scale terms, $\sigma_i$.
In fact, you can think of homoskedasticity as the simplest such model; assuming that all $\sigma_i = \sigma$.
A more general model of $\sigma_i$ should encode any information the analyst has about the scale terms.
This can be a distribution or functions of covariates for how we think observations may have different values.

### Covariates

A simple model of heteroskedasticity is if the observations can be split into groups. Suppose the observations are partitioned into $k = 1, \dots, K$ groups, and $k[i]$ is the group of observation $i$,
$$
\sigma_i = \sigma_{k[i]}
$$

Another choice would be to model the scale term with a regression model, for example,
$$
\log(\sigma_i) \sim \dnorm(X \gamma, \tau)
$$


### Student-t

It turns out that the Student-t distribution of error terms from the [Robust Regression] chapter can also be derived as a model of heteroskedasticity.

A reparameterization that will be used quite often is to rewrite a normal distributions with unequal
scale parameters as a continous mixture of a common global scale parameter ($\sigma$), and observation specific local scale parameters, $\lambda_i$,[^globalmixture]
$$
y_i \sim \dnorm(X\beta, \lambda_i \sigma) .
$$

If the local scale paramters are distributed as,
$$
\lamba^2 \sim \dinvgamma(\nu / 2, \nu / 2)
$$
then the above is equivalent to a regression with errors distributed Student-t errors with $\nu$ degrees of freedom,
$$
y_i \sim \dt{\nu}(X \beta, \sigma) .
$$

[^globalmixture] See [this](http://www.sumsar.net/blog/2013/12/t-as-a-mixture-of-normals/) for a visualization of a Student-t distribution a mixture of Normal distributions, and [this](https://www.johndcook.com/t_normal_mixture.pdf) for a derivation of the Student t distribution as a mixture of normals. This scale mixture of normals representation will also be used with shrinkage priors on the regression coefficients.


**Example:** Simulate Student-t distribution with $\nu$ degrees of freedom as a scale mixture of normals. For *s in 1:S$,

1. Simulate $z_s \sim \dgamma(\nu / 2, \nu / 2)$
2. $x_s = 1 / \sqrt{z_s}2$ is draw from $\dt{\nu}(0, 1)$.

When using R, ensure that you are using the correct parameterization of the gamma distribution. **Left to reader**


## References

### Robust regression 

- See @GelmanHill2008a [sec 6.6], @BDA2013 [ch 17]
- @Stan2016a [Sec 8.4] for the Stan example using a Student-t distribution

### Heteroskedasticity

- @BDA201 [Sec. 14.7] for models with unequal variances and correlations.
- @Stan2016a reparameterizes the Student t distribution as a mixture of gamma distributions in Stan.
