```{r}
library("datums")
library("tidyverse")
```

Frank Weldon threw 12 dice 26,306 times and counted the number of dice that showed a five or six.

```{r}
N <- 26306
prob <- 2 / 6
size <- 12
```

For a prior, let's suppose that one roll of 12 dice has 4 dice showing 5's or 6's and 8 dice showing other numbers.
This is equivalent of $\mathrm{Beta}(4, 8)$.
```{r}
prior_a <- 4
prior_b <- 8
```

The total number of 5's and 6's in the $`r N` \times 12$ dice rolls.
```{r}
total <- sum(weldon[["observed"]] * weldon[["n"]])
```

The posterior distribution is,
```{r}
posterior_a <- prior_a + total
posterior_b <- prior_b + (N * size - total)
```

The posterior distribution is:
```{r}
qbeta(c(0.025, 0.975), posterior_a, posterior_b)
```
The posterior expected value is
```{r}
posterior_a / (posterior_a + posterior_b)
```
The mode is
```{r}
(posterior_a - 1) / (posterior_a + posterior_b - 2)
```
The median is approximately
```{r}
(posterior_a - 1 / 3) / (posterior_a + posterior_b - 2 / 3)
```

The probability that the observed data is greater than the theoretical quantile:
```{r}
pbeta(prob, posterior_a, posterior_b, lower.tail = FALSE)
```

Plotting the prior and the posterior on the data:
```{r}
tibble(x = ppoints(100),
       posterior = dbeta(x, posterior_a, posterior_b),
       prior = dbeta(x, prior_a, prior_b)) %>%
  gather(variable, density, -x) %>%
  ggplot(aes(x = x, y = density, colour = variable)) +
  geom_line()
```

```{r}

```

Alternatively, we could instead model the proportion as a multinomial distribution.
This treats the categories as 1, ..., 12, as independent events.

The prior distribution is Dirichlet with parameters derived from the expected number in each category for a single roll of `r size` dice.
```{r}
prior_alpha <- dbinom(0:size, size = size, prob = prob) * size
round(prior_alpha , 1)
```

The posterior distribution for each category is also a Dirichlet distribution with the following parameters,
```{r}
post_alpha <- prior_alpha + weldon[["observed"]]
round(post_alpha, 1)
```

The posterior predictive distribution is the [Dirichlet-multinomial](https://en.wikipedia.org/wiki/Dirichlet-multinomial_distribution) distribution.


```{r}

```


